Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-11 paper.qmd
Update the color of the points and consistant with coloring

data <- read_csv(paste0(here::here(), "/data/s_curve.csv"))
```

@@

plot_list1 + plot_list2 + plot_list3 + plot_list4 + plot_list5 +
  plot_layout(ncol=5, widths=c(0.35, 0.15, 0.30, 0.12, 0.11))
```

@@

### Total number of bins

@@

### Benchmark value to remove low-density hexagons

@@

### Benchmark value to remove long edges

@@

### Summaries of the model {#sec-summary}

@@

The concept of "residuals" is pivotal in evaluating the accuracy of representing bin centroids in high dimensions. To quantify this accuracy, we introduce an error metric, which measures the sum of squared differences between the high-dimensional data (\(x_{ij}\)) and the predicted bin centroid data in high-dimensional space (\(C_{x_ij}\)) across all bins and dimensions. Mathematically, this error is expressed as:

@@

Here, \(n\) represents the number of bins, \(p\) represents the dimensions, \(x_{ij}\) is the actual high-dimensional data, and \(C_{x_ij}\) is the predicted bin centroid data in high dimensions.

@@

data <- read_csv("data/s_curve.csv")
data <- data |>
  mutate(ID = row_number())

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_discrete_qualitative() +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df, aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_discrete_qualitative() +
  ylab("MSE") +
  xlab("Total number of bins")

@@

@@

MSE_df_umap <- dplyr::bind_rows(MSE_df_umap_1, MSE_df_umap_2, MSE_df_umap_3, MSE_df_umap_4)

## To draw with AIC
aic_plot_param <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                                         y = total_error,
                                                                         color = param
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_discrete_qualitative() +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot_param_training <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                               y = total_mse,
                               color = param
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_discrete_qualitative() +
  ylab("MSE") +
  xlab("Total number of bins")

mse_plot_param_test <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "test"), aes(x = number_of_bins,
                                                                                             y = total_mse,
                                                                                             color = param
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_discrete_qualitative() +
  ylab("MSE") +
  xlab("Total number of bins")

@@

## To draw with AIC
aic_gau_plot <- ggplot(MSE_df_gau |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                             y = total_error,
                                                             color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_discrete_qualitative() +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_gau_plot <- ggplot(MSE_df_gau, aes(x = number_of_bins,
                   y = total_mse,
                   color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_discrete_qualitative() +
  ylab("MSE") +
  xlab("Total number of bins")

@@

<!--six different clusters-->

<!--Gaussian globe or sphere-->

<!--In this section, we demonstrate the effectiveness of our approach through the use of simulation data. These simulations are designed to have known cluster structures and underlying data geometries. We begin with a straightforward example where we generate spherical clusters with a doublet structure. This allows us to assess how well our algorithm can identify and represent these known structures in the low-dimensional space.

We then proceed to a more complex example where the data exhibits a more intricate geometry. This simulation provides a challenging test for our approach to capture and preserve the underlying data structure accurately.

**Example 1: Exploring spherical Gaussian clusters with doublet structure**

The next data set consists of two distinct types of clusters embedded in a 10-$d$ space. The first set of clusters includes three Gaussian clusters with equal variance. Each of these Gaussian clusters is equidistant from the others, forming a well-separated and symmetrical arrangement. The second set of clusters consists of three clusters with a doublet structure. These clusters are also embedded in the 10-$d$ space, but each is positioned in the middle of its parent Gaussian clusters. Moreover, each doublet cluster contains an average number of points from its corresponding parent Gaussian clusters.

For analysis, we run tSNE on the training data with a perplexity setting 18. @fig-example3 (a) displays the tSNE results, indicating that the sub-clusters have been correctly identified. However, the relative locations of certain clusters to each other appear distorted in the low-dimensional view, such as the red and pink clusters being far apart.

Nonetheless, we find no apparent impact on the perception of the data structure when we examine the tour view and the model (see @fig-example3) through the linked video (see @fig-exp1_sc). The model demonstrates accurate identification of clusters and captures the internal variations within each cluster, reflecting the inherent variance present in the data. The configurations used for the analysis are detailed in @tbl-table02. Additionally, the representation of the parent clusters in the low-dimensional space appears stretched, aligning with the variance of these clusters. On the other hand, the doublet clusters are squeezed, reflecting their specific characteristics and relationship to their parent Gaussian clusters.

```{r}
#| warning: false
#| echo: false
sample_size <- 450
cluster_size <- sample_size/4.5

df1 <- tibble::tibble(x1=rnorm(cluster_size, mean = 3, sd = 0.05), x2 = rnorm(cluster_size, mean = 1, sd = 0.05), x3=rnorm(cluster_size, mean = 1, sd = 0.05), x4=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x5=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x6=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x7=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x8=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x9=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x10=rnorm(cluster_size, mean = 1, sd = 0.05))

df1 <- df1 %>%
  mutate(cluster = 1)

df2 <- tibble::tibble(x1=rnorm(cluster_size, mean = 1, sd = 0.05), x2=rnorm(cluster_size, mean = 1, sd = 0.05), x3=rnorm(cluster_size, mean = 1, sd = 0.05), x4=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x5=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x6=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x7=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x8=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x9=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x10=rnorm(cluster_size, mean = 1, sd = 0.05))

df2 <- df2 %>%
  mutate(cluster = 2)

df3_new <- (df1 + df2) / 2

samp <- sample(nrow(df3_new), cluster_size * 0.50) ## 50% from the original dataset

# data in the sample
df3 <- df3_new[samp,]

df3 <- df3 %>%
  mutate(cluster = 3)

df4 <- tibble::tibble(x1=rnorm(cluster_size, mean = 1, sd = 0.05), x2=rnorm(cluster_size, mean = 1, sd = 0.05), x3=rnorm(cluster_size, mean = 1, sd = 0.05), x4=rnorm(cluster_size, mean = 3, sd = 0.05),
                      x5=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x6=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x7=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x8=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x9=rnorm(cluster_size, mean = 1, sd = 0.05),
                      x10=rnorm(cluster_size, mean = 1, sd = 0.05))

df4 <- df4 %>%
  mutate(cluster = 4)

df5_new <- (df2 + df4) / 2

samp1 <- sample(nrow(df5_new), cluster_size * 0.50) ## 50% from the original dataset

df5 <- df5_new[samp1,]

df5 <- df5 %>%
  mutate(cluster = 5)

df6_new <- (df1 + df4) / 2

samp2 <- sample(nrow(df6_new), cluster_size * 0.50) ## 50% from the original dataset

df6 <- df6_new[samp2,]

df6 <- df6 %>%
  mutate(cluster = 6)

df_3 <- bind_rows(df1, df2, df3, df4, df5, df6)
```

# langevitour::langevitour(df_3[, 1:10], group = df_3$cluster, levelColors = c("#b15928", "#1f78b4", "#bc80bd", "#33a02c", "#fb9a99", "#e31a1c"))
```

df_3 <- df_3 |>
  mutate(ID = row_number())

data_split_sp <- initial_split(df_3)
training_data_5 <- training(data_split_sp) |>
  arrange(ID)
test_data_5 <- testing(data_split_sp) |>
  arrange(ID)

clusters <- training_data_5$cluster
```

```{r}
#| warning: false
#| echo: false 
tSNE_data_gau <- Fit_tSNE(training_data_5 |> dplyr::select(-ID, -cluster), with_seed = 20240110)
```

tSNE_cluster <- tSNE_data_gau %>%
  ggplot(aes(x = tSNE1,
             y = tSNE2, color = as.character(clusters))) +
  geom_point(alpha=0.5) +
  coord_equal() +
  theme_linedraw() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 7), #change legend key width
        legend.title = element_text(size=7), #change legend title font size
        legend.text = element_text(size=7)) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#bc80bd", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "Cluster"))

cell_area <- 1

num_bins_tsne_gau <- calculate_effective_x_bins(.data = tSNE_data_gau, x = tSNE1,
                                                cell_area = 1)
shape_val_tsne_gau <- calculate_effective_shape_value(.data = tSNE_data_gau,
                                                      x = tSNE1, y = tSNE2)

## To extract bin centroids
hexbin_data_object_tsne_gau <- extract_hexbin_centroids(nldr_df = tSNE_data_gau, num_bins = num_bins_tsne_gau, shape_val = shape_val_tsne_gau, x = tSNE1, y = tSNE2)

df_bin_centroids_tsne_gau <- hexbin_data_object_tsne_gau$hexdf_data

# min_std_cell_threshold3 <- 0.01
# 
# df_bin_centroids_pbmc <- df_bin_centroids_pbmc |>
#   dplyr::mutate(stand_cell_count = counts/max(counts)) |>
#   dplyr::filter(stand_cell_count > min_std_cell_threshold3)

tSNE_data_with_hb_id_gau <- tSNE_data_gau |> 
  dplyr::mutate(hb_id = hexbin_data_object_tsne_gau$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all_tsne_gau <- dplyr::bind_cols(training_data_5 |> dplyr::select(-ID), tSNE_data_with_hb_id_gau)

## Averaged on high-D
df_bin_tsne_gau <- avg_highD_data(.data = df_all_tsne_gau)

## Triangulate bin centroids
tr1_object_tsne_gau <- triangulate_bin_centroids(df_bin_centroids_tsne_gau, x, y)
tr_from_to_df_tsne_gau <- generate_edge_info(triangular_object = tr1_object_tsne_gau)

## Compute 2D distances
distance_tsne_gau <- cal_2D_dist(.data = tr_from_to_df_tsne_gau)

## To find the benchmark value
benchmark_tsne_gau <- find_benchmark_value(.data = distance_tsne_gau, distance_col = distance)
#benchmark_tsne <- 5

tour_tsne_gau <- show_langevitour(df_all_tsne_gau, df_bin_tsne_gau, df_bin_centroids_tsne_gau, benchmark_value = benchmark_tsne_gau, distance = distance_tsne_gau, distance_col = distance)

trimesh_cluster <- ggplot(df_bin_centroids_tsne_gau, aes(x = x, y = y)) +
  geom_trimesh() +
  coord_equal()

trimesh_cluster <- trimesh_cluster +
  #ggtitle("(b)") +
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))

trimesh_gr_cluster <- colour_long_edges(.data = distance_tsne_gau, benchmark_value = benchmark_tsne_gau,
                                     triangular_object = tr1_object_tsne_gau, distance_col = distance)

trimesh_gr_cluster <- trimesh_gr_cluster +
  #ggtitle("(c)") +
  xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7), #change legend key width
        legend.title = element_text(size=7), #change legend title font size
        legend.text = element_text(size=5))

tour_cluster <- show_langevitour(df_all_tsne_gau, df_bin_tsne_gau, df_bin_centroids_tsne_gau, benchmark_value = benchmark_tsne_gau, distance = distance_tsne_gau, distance_col = distance)
```

```{r}
#| warning: false
#| echo: false
#| out-width: 100%
#| label: fig-example3
#| fig-cap: Visualization and refinement of low-dimensional representation for sub-clustered data$\colon$ (a) 2D layout of tSNE (perplexity = 18), (b) triangular mesh, and (c) triangular mesh colored by edge type. Edges with lengths less than the benchmark value of $2$ are classifiedas small, while edges with lengths greater than or equal to this value are labeled as long. The combination of tSNE, hexagonal binning, and triangulation provides the effectiveness of our approach in visualizing and refining the low-dimensional representation of the sub-clustered data.

tSNE_cluster + trimesh_cluster + trimesh_gr_cluster +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect') &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

::: {#fig-exp1_sc layout-ncol="3"}
![](figures/example_1a.png){#fig-exp1_sc1}

![](figures/example_1b.png){#fig-exp1_sc2}

![](figures/example_1c.png){#fig-exp1_sc3}

Screen shots of the **langevitour** with the low-dimensional view applied to sub-clustered data set, a video of the tour animation is available at (\url{https://drive.google.com/file/d/15nAbmc4UoVlUyFcGtnlnmyxrJEbK39nx/view}). Combination of the **langevitour** and low-dimensional representation of the sub-clustered data provides the effectiveness of our approach.
:::

**Example 2: Exploring data with piecewise linear structure**

In our next exploration, we delve into simulated noisy tree-structured data designed to mimic branching trajectories of cell differentiation. The simulation is crafted to resemble the hierarchical clustering pattern observed in cell differentiation trajectories, where mature cells are akin to the tips of branches.

To begin our analysis, we apply principal components to the data and focus on the first ten principal components, collectively accounting for approximately 68% of the variance explained in the data set. Then, we run TriMAP with the default parameters on the training data (n_inliers: 5, n_outliers: 4, and n_random: 3).  Through this process, we create a low-dimensional manifold using the configurations detailed in @tbl-table02. Finally, we present the model along with the data using the tour view (Figure @fig-exp2_sc), allowing us to visualize and explore the intricate relationships and structures present in the data's low-dimensional representation.

data_phate <- read_csv(paste0(here::here(), "/data/data_phate.csv"))

data_phate <- data_phate |>
  dplyr::mutate(ID = row_number())
```

```{r}
#| warning: false
#| echo: false 
data_split <- initial_split(data_phate)
training_data <- training(data_split) |>
  arrange(ID)
test_data <- testing(data_split) |>
  arrange(ID)
```

```{r}
#| warning: false
#| echo: false 
## num_pcs: selected number of PCs to get results (deafukt is all)

calculate_pca <- function(data, num_pcs = NCOL(data), center = TRUE, scale = FALSE){
  pcaY_cal <- prcomp(data, center = TRUE, scale = FALSE)

  PCAresults <- tibble::as_tibble(pcaY_cal$x[, 1:num_pcs])
  names(PCAresults) <- paste0(rep("PC", num_pcs), 1:num_pcs)

  summary_pca <- summary(pcaY_cal)

  var_explained_df <- tibble::tibble(PC = paste0("PC",1:NCOL(data)),
                               var_explained = (pcaY_cal$sdev[1:NCOL(data)])^2/sum((pcaY_cal$sdev[1:NCOL(data)])^2))

  ## Scree plot
  plot <- var_explained_df %>%
    ggplot(aes(x = PC,y = var_explained, group = 1))+
    geom_point(size=1)+
    geom_line() +
    theme_minimal() +
    scale_x_discrete(limits = paste0(rep("PC", 20), 1:20)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.text = element_text(size = 5),
          axis.title = element_text(size = 7)) +
    xlab("Principal Components (PC)") +
    ylab("Variance explained")

  return(list(prcomp_out = pcaY_cal, pca_components = PCAresults, summary = summary_pca, var_explained_pca  = var_explained_df, plot = plot))
}
```

```{r}
#| warning: false
#| echo: false 
pca_out <- calculate_pca(training_data, 10)
data_pca <- pca_out$pca_components
data_pca <- data_pca %>% 
  distinct()

var_explain_example2 <- pca_out$plot
```

```{r}
#| warning: false
#| echo: false 
TriMAP_data_tree <- read_csv(paste0(here::here(), "/data/trimap_data_linear_structure_pca.csv")) %>%
    mutate(ID = row_number())

```{r}
#| warning: false
#| echo: false 
TriMAP_tree <- plot_TriMAP_2D(TriMAP_data_tree) +
  theme_linedraw() +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

cell_area <- 1

num_bins_tree <- calculate_effective_x_bins(.data = TriMAP_data_tree, x = TriMAP1,
                                            cell_area = 1)
shape_val_tree <- calculate_effective_shape_value(.data = TriMAP_data_tree,
                                                  x = TriMAP1, y = TriMAP2)

## To extract bin centroids
hexbin_data_object_tree <- extract_hexbin_centroids(nldr_df = TriMAP_data_tree, num_bins = num_bins_tree, shape_val = shape_val_tree, x = TriMAP1, y = TriMAP2)

df_bin_centroids_tree <- hexbin_data_object_tree$hexdf_data

# min_std_cell_threshold3 <- 0.01
# 
# df_bin_centroids_tree <- df_bin_centroids_tree |>
#   dplyr::mutate(stand_cell_count = counts/max(counts)) |>
#   dplyr::filter(stand_cell_count > min_std_cell_threshold3)

TriMAP_data_with_hb_id_tree <- TriMAP_data_tree |> 
  dplyr::mutate(hb_id = hexbin_data_object_tree$hb_data@cID)

## To generate a data set with high-D and 2D training data
data_pca_n <- data_pca
names(data_pca_n) <- paste0(rep("x", 7), 1:7)
df_all_tree <- dplyr::bind_cols(data_pca_n, TriMAP_data_with_hb_id_tree)

## Averaged on high-D
df_bin_tree <- avg_highD_data(.data = df_all_tree)

## Triangulate bin centroids
tr1_object_tree <- triangulate_bin_centroids(df_bin_centroids_tree, x, y)
tr_from_to_df_tree <- generate_edge_info(triangular_object = tr1_object_tree)

>>>>

data <- read_rds("data/s_curve/s_curve.rds")
```

@@

plot_list1 + plot_list2 + plot_list3 + plot_list4 + plot_list5 +
  plot_layout(ncol=5)
```

@@

#### Total number of bins

@@

#### Benchmark value to remove low-density hexagons

@@

#### Benchmark value to remove long edges

@@

### Model summaries {#sec-summary}

@@

The concept of "residuals" is pivotal in evaluating the accuracy of representing bin centroids in high dimensions. To quantify this accuracy, we introduce an error metric, which measures the sum of squared differences between the high-dimensional data ($x_{ij}$) and the predicted bin centroid data in high-dimensional space ($C_{x_ij}$) across all bins and dimensions. Mathematically, this error is expressed as:

@@

Here, $n$ represents the number of bins, $p$ represents the dimensions, $x_{ij}$ is the actual high-dimensional data, and $C_{x_ij}$ is the predicted bin centroid data in high dimensions.

@@

data <- read_rds("data/s_curve/s_curve.rds")

@@

MSE_df$method <- factor(MSE_df$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df, aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Total number of bins")

@@

<!--UMAP with different param choices-->

UMAP_data_7 <- read_rds(file = "data/s_curve/s_curve_umap_7.rds")

#(n-neighbors: 50)
plot_list1_umap <- plot_UMAP_2D(UMAP_data_7) + #ggtitle("(b)") +
  geom_point(color = "#8dd3c7") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

UMAP_data_15 <- read_rds(file = "data/s_curve/s_curve_umap_15.rds")

#(n-neighbors: 50)
plot_list2_umap <- plot_UMAP_2D(UMAP_data_15) + #ggtitle("(b)") + 
  geom_point(color = "#a65628") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

UMAP_data_32 <- read_rds(file = "data/s_curve/s_curve_umap_32.rds")

#(n-neighbors: 50)
plot_list3_umap <- plot_UMAP_2D(UMAP_data_32) + #ggtitle("(b)") +
  geom_point(color = "#f781bf") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

#(n-neighbors: 50)
plot_list4_umap <- plot_UMAP_2D(UMAP_data) + #ggtitle("(b)") +
  geom_point(color = "#999999") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "2D layouts from UMAP applied for the S-curve data: (a) UMAP (n_neighbors = 7), (b) UMAP (n_neighbors = 15), (c) UMAP (n_neighbors = 32), (d) UMAP (n_neighbors = 50). Is there a best parameter choice in representing UMAP or are they all providing  equivalent information?"
#| label: fig-nldervisUMAP
#| out-width: 100%

plot_list1_umap + plot_list2_umap + plot_list3_umap + plot_list4_umap +
  plot_layout(ncol=4)
```

@@

MSE_df_umap <- dplyr::bind_rows(MSE_df_umap_1, MSE_df_umap_2, MSE_df_umap_3, MSE_df_umap_4)
MSE_df_umap$param <- factor(MSE_df_umap$param, levels = c("n_neighbors: 7", "n_neighbors: 15", "n_neighbors: 32", "n_neighbors: 50"))

## To draw with AIC
aic_plot_param <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                                         y = total_error,
                                                                         color = param
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_color_manual(values = c("#8dd3c7", "#a65628", "#f781bf", "#999999")) +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot_param_training <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                               y = total_mse,
                               color = param
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_manual(values = c("#8dd3c7", "#a65628", "#f781bf", "#999999")) +
  ylab("MSE") +
  xlab("Total number of bins")

mse_plot_param_test <- ggplot(MSE_df_umap |> dplyr::filter(data_type == "test"), aes(x = number_of_bins,
                                                                                             y = total_mse,
                                                                                             color = param
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), legend.text = element_text(size=7), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_color_manual(values = c("#8dd3c7", "#a65628", "#f781bf", "#999999")) +
  ylab("MSE") +
  xlab("Total number of bins")

@@

MSE_df_gau$method <- factor(MSE_df_gau$method, levels = c("tSNE", "UMAP", "PHATE", "TriMAP", "PaCMAP"))

## To draw with AIC
aic_gau_plot <- ggplot(MSE_df_gau |> dplyr::filter(data_type == "training"), aes(x = number_of_bins,
                                                             y = total_error,
                                                             color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_gau_plot <- ggplot(MSE_df_gau, aes(x = number_of_bins,
                   y = total_mse,
                   color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Total number of bins")

@@
